---
title: "Task 6: Genome-Wide Association Study (GWAS) Analysis"
subtitle: "Severe COVID-19 with Respiratory Failure — Ellinghaus et al. 2020"
author: "ghania | 460673"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 12,
  fig.height = 7,
  dpi        = 150
)
```

---

# 1. Dataset Source and Description

| Field               | Details |
|---------------------|---------|
| **GWAS Catalog ID** | GCST90000255 |
| **Study**           | Ellinghaus D et al. (2020), *N Engl J Med* |
| **Title**           | Genomewide Association Study of Severe Covid-19 with Respiratory Failure |
| **Trait**           | Severe COVID-19 infection with respiratory failure (Analysis I) |
| **Population**      | 3,815 European ancestry individuals |
| **Genome Build**    | GRCh38 |
| **File**            | `GCST90000255_GRCh38.tsv.gz` |
| **Source**          | NHGRI-EBI GWAS Catalog |

---

# 2. Load Required Packages

```{r load-packages}
library(data.table)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(scales)
library(RColorBrewer)
library(knitr)
```

---

# 3. Load Data

```{r load-data}
gwas_file <- "GCST90000255_GRCh38.tsv.gz"

cat("Reading GWAS summary statistics...\n")
gwas_raw <- fread(gwas_file, sep = "\t", header = TRUE)

cat("Dimensions:", nrow(gwas_raw), "rows x", ncol(gwas_raw), "columns\n")
cat("Original column names:", paste(names(gwas_raw), collapse = ", "), "\n")

head(gwas_raw, 3)
```

---

# 4. Preprocessing

```{r preprocess}
# Rename columns directly using known original names
gwas <- gwas_raw
names(gwas) <- c("CHR", "BP", "SNP", "EA", "NEA", "BETA", "SE", "P")

# Convert types and filter to autosomes with valid p-values
gwas <- gwas %>%
  mutate(
    CHR  = as.integer(CHR),
    BP   = as.integer(BP),
    P    = as.numeric(P),
    MAF  = NA_real_
  ) %>%
  filter(
    !is.na(CHR), !is.na(BP), !is.na(P),
    CHR %in% 1:22,
    P > 0, P <= 1
  ) %>%
  mutate(
    NEGLOG10P   = -log10(P),
    SIGNIFICANT = P < 5e-8
  )

cat("SNPs after QC filters:", nrow(gwas), "\n")
cat("Genome-wide significant SNPs (P < 5e-8):", sum(gwas$SIGNIFICANT), "\n")
cat("Suggestive significant SNPs (P < 1e-5):",  sum(gwas$P < 1e-5), "\n")

# Save raw SNP count before freeing RAM (used in QC table below)
n_raw_snps <- nrow(gwas_raw)

# Free RAM — gwas_raw no longer needed
rm(gwas_raw)
gc()
```

---

# 5. Quality Control Summary

```{r qc-table}
qc_df <- data.frame(
  Step  = c("Raw SNPs loaded",
            "After autosome + valid P filter",
            "Genome-wide significant (P < 5e-8)",
            "Suggestive significant (P < 1e-5)"),
  Count = c(n_raw_snps,
            nrow(gwas),
            sum(gwas$SIGNIFICANT),
            sum(gwas$P < 1e-5))
)
knitr::kable(qc_df, caption = "QC Step Summary")
```

---

# 6. Principal Component Analysis (PCA)

> **Note:** Standard GWAS PCA requires individual-level genotype data to detect
> population stratification. Since GCST90000255 provides only summary statistics,
> we perform a statistics-level PCA on SNP features (log10P, BETA, SE) to
> illustrate the PCA concept in R. The original Ellinghaus et al. study performed
> genotype PCA and confirmed European ancestry homogeneity before association testing.

```{r pca}
set.seed(42)

pca_input <- gwas %>%
  select(CHR, NEGLOG10P, BETA, SE) %>%
  filter(complete.cases(.)) %>%
  slice_sample(n = min(50000, nrow(.)))

pca_mat    <- scale(pca_input %>% select(-CHR))
pca_result <- prcomp(pca_mat, center = FALSE, scale. = FALSE)

pca_df <- as.data.frame(pca_result$x[, 1:3]) %>%
  mutate(CHR = factor(pca_input$CHR))

var_explained <- round(100 * pca_result$sdev^2 / sum(pca_result$sdev^2), 2)

cat("Variance explained — PC1:", var_explained[1], "%  PC2:", var_explained[2], "%\n")
```

```{r pca-plot, fig.cap="PCA of SNP-level statistics coloured by chromosome"}
ggplot(pca_df, aes(x = PC1, y = PC2, colour = CHR)) +
  geom_point(alpha = 0.4, size = 0.7) +
  scale_colour_manual(
    values = rep(c("#1f77b4","#ff7f0e","#2ca02c","#d62728","#9467bd",
                   "#8c564b","#e377c2","#7f7f7f","#bcbd22","#17becf",
                   "#aec7e8","#ffbb78"), length.out = 22)
  ) +
  labs(
    title    = "PCA of GWAS Summary Statistics",
    subtitle = paste0("PC1: ", var_explained[1], "%  |  PC2: ", var_explained[2], "% variance explained"),
    x        = paste0("PC1 (", var_explained[1], "%)"),
    y        = paste0("PC2 (", var_explained[2], "%)"),
    colour   = "Chromosome",
    caption  = "Based on -log10(P), BETA, SE for a random sample of 50,000 SNPs"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.key.size = unit(0.4, "cm"),
        legend.text     = element_text(size = 7))

ggsave("PCA_Plot.png", width = 12, height = 7, dpi = 200)
cat("Saved: PCA_Plot.png\n")
```

```{r scree-plot, fig.width=7, fig.height=5, fig.cap="Scree plot — variance explained per PC"}
scree_df <- data.frame(
  PC  = 1:min(10, length(var_explained)),
  Var = var_explained[1:min(10, length(var_explained))]
)

ggplot(scree_df, aes(x = PC, y = Var)) +
  geom_col(fill = "#2c7bb6", colour = "white") +
  geom_line(colour = "red", linewidth = 0.8) +
  geom_point(colour = "red", size = 2.5) +
  scale_x_continuous(breaks = scree_df$PC) +
  labs(title = "Scree Plot — Variance Explained by Each PC",
       x = "Principal Component", y = "Variance Explained (%)") +
  theme_bw(base_size = 13)
```

---

# 7. Manhattan Plot

```{r manhattan-prep}
# Thin data to reduce RAM — keep all significant SNPs + P < 0.05 subset
gwas <- gwas %>% filter(P < 0.05 | SIGNIFICANT)

# Compute cumulative base-pair positions per chromosome
chr_lengths <- gwas %>%
  group_by(CHR) %>%
  summarise(chr_len = max(BP), .groups = "drop") %>%
  arrange(CHR) %>%
  mutate(tot = lag(cumsum(chr_len), default = 0))

gwas_plot <- gwas %>%
  left_join(chr_lengths, by = "CHR") %>%
  mutate(BPcum = BP + tot)

# X-axis tick at midpoint of each chromosome
axisdf <- gwas_plot %>%
  group_by(CHR) %>%
  summarise(centre = mean(BPcum), .groups = "drop")

# Lead SNP per chromosome for labelling
sig_snps <- gwas_plot %>%
  filter(SIGNIFICANT) %>%
  group_by(CHR) %>%
  slice_min(P, n = 1) %>%
  ungroup()

cat("Significant loci to label:", nrow(sig_snps), "\n")
```

```{r manhattan-plot, fig.width=14, fig.height=6, fig.cap="Manhattan plot — Severe COVID-19 GWAS"}
chr_colours <- rep(c("#2c7bb6", "#abd9e9"), 11)

p_manhattan <- ggplot(gwas_plot,
                      aes(x = BPcum, y = NEGLOG10P, colour = factor(CHR))) +
  geom_point(alpha = 0.5, size = 0.6) +
  geom_point(data = gwas_plot %>% filter(SIGNIFICANT),
             colour = "#d73027", size = 1.8, alpha = 0.9) +
  geom_hline(yintercept = -log10(5e-8), linetype = "dashed",
             colour = "red", linewidth = 0.7) +
  geom_hline(yintercept = -log10(1e-5), linetype = "dotted",
             colour = "blue", linewidth = 0.5) +
  geom_label_repel(
    data           = sig_snps,
    aes(label      = SNP),
    colour         = "black",
    fill           = "white",
    size           = 2.8,
    box.padding    = 0.4,
    max.overlaps   = 20,
    segment.colour = "grey40"
  ) +
  scale_colour_manual(values = chr_colours, guide = "none") +
  scale_x_continuous(label = axisdf$CHR, breaks = axisdf$centre) +
  scale_y_continuous(
    expand = c(0, 0),
    limits = c(0, max(gwas_plot$NEGLOG10P, na.rm = TRUE) * 1.1)
  ) +
  labs(
    title    = "Manhattan Plot — Severe COVID-19 GWAS",
    subtitle = "Ellinghaus et al. 2020  |  GCST90000255  |  GRCh38",
    x        = "Chromosome",
    y        = expression(-log[10](italic(p))),
    caption  = "Red dashed: genome-wide significance (p = 5x10-8)   Blue dotted: suggestive (p = 1x10-5)"
  ) +
  theme_bw(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        axis.text.x        = element_text(size = 9))

print(p_manhattan)
ggsave("Manhattan_Plot.png", p_manhattan, width = 14, height = 6, dpi = 200)
cat("Saved: Manhattan_Plot.png\n")
```

---

# 8. QQ Plot

```{r qq-plot, fig.width=7, fig.height=7, fig.cap="QQ plot of observed vs expected p-values"}
n_snps   <- nrow(gwas)
observed <- sort(gwas$NEGLOG10P, decreasing = TRUE)
expected <- -log10(ppoints(n_snps))

# Genomic inflation factor lambda
chisq  <- qchisq(gwas$P, df = 1, lower.tail = FALSE)
lambda <- round(median(chisq, na.rm = TRUE) / qchisq(0.5, df = 1), 4)
cat("Genomic inflation factor lambda =", lambda, "\n")

# 95% confidence band
ci_upper <- -log10(qbeta(0.025, 1:n_snps, n_snps:1))
ci_lower <- -log10(qbeta(0.975, 1:n_snps, n_snps:1))

qq_df <- data.frame(expected = expected, observed = observed,
                    upper = ci_upper, lower = ci_lower)

p_qq <- ggplot(qq_df, aes(x = expected, y = observed)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "lightblue", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, colour = "red", linewidth = 0.8) +
  geom_point(colour = "#2c7bb6", size = 0.6, alpha = 0.6) +
  labs(
    title    = "QQ Plot — Severe COVID-19 GWAS",
    subtitle = paste0("Genomic inflation factor  lambda = ", lambda),
    x        = expression("Expected " * -log[10](italic(p))),
    y        = expression("Observed " * -log[10](italic(p))),
    caption  = "Light blue = 95% CI under null hypothesis\nlambda close to 1.0 indicates well-controlled inflation"
  ) +
  theme_bw(base_size = 13)

print(p_qq)
ggsave("QQ_Plot.png", p_qq, width = 7, height = 7, dpi = 200)
cat("Saved: QQ_Plot.png\n")
```

---

# 9. Significant SNP Tables

```{r sig-snps-table}
sig_table <- gwas %>%
  filter(P < 5e-8) %>%
  arrange(P) %>%
  select(CHR, BP, SNP, EA, NEA, BETA, SE, P, NEGLOG10P) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4)))

cat("Genome-wide significant SNPs:", nrow(sig_table), "\n\n")
knitr::kable(head(sig_table, 30),
             caption = "Top Genome-Wide Significant SNPs (P < 5x10-8)")
```

```{r suggestive-table}
sug_table <- gwas %>%
  filter(P >= 5e-8 & P < 1e-5) %>%
  arrange(P) %>%
  select(CHR, BP, SNP, EA, NEA, BETA, SE, P, NEGLOG10P) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4)))

cat("Suggestive significant SNPs:", nrow(sug_table), "\n\n")
knitr::kable(head(sug_table, 20),
             caption = "Suggestive Significant SNPs (P < 1x10-5)")
```

---

# 10. Per-Chromosome Summary

```{r chr-summary}
chr_summary <- gwas %>%
  group_by(CHR) %>%
  summarise(
    Total_SNPs      = n(),
    Min_P           = signif(min(P), 3),
    Max_NegLog10P   = round(max(NEGLOG10P), 2),
    GW_Sig_SNPs     = sum(P < 5e-8),
    Suggestive_SNPs = sum(P < 1e-5),
    .groups = "drop"
  ) %>%
  arrange(CHR)

knitr::kable(chr_summary, caption = "Per-Chromosome GWAS Summary")
```

---

# 11. Save All Results

```{r save-results}
fwrite(sig_table,   "GWAS_GW_significant_SNPs.csv")
fwrite(sug_table,   "GWAS_suggestive_SNPs.csv")
fwrite(chr_summary, "GWAS_chromosome_summary.csv")
fwrite(gwas %>% arrange(P) %>%
         select(CHR, BP, SNP, EA, NEA, BETA, SE, P, NEGLOG10P, SIGNIFICANT),
       "GWAS_all_cleaned.csv")

cat("All files saved:\n")
cat("  GWAS_GW_significant_SNPs.csv\n")
cat("  GWAS_suggestive_SNPs.csv\n")
cat("  GWAS_chromosome_summary.csv\n")
cat("  GWAS_all_cleaned.csv\n")
cat("  PCA_Plot.png\n")
cat("  Manhattan_Plot.png\n")
cat("  QQ_Plot.png\n")
```

---

# 12. Biological Interpretation

## Key GWAS Loci Identified

| Locus | Chromosome | Key Gene(s) | Biological Role |
|-------|-----------|-------------|-----------------|
| **3p21.31** | Chr 3 | *SLC6A20, LZTFL1, CCR9, FYCO1, CXCR6, XCR1* | Immune trafficking, ACE2 interaction |
| **9q34.2**  | Chr 9 | *ABO* blood group locus | Endothelial function, coagulation |

## Interpretation

- **Chr 3 locus (3p21.31):** The strongest signal in this study. *SLC6A20* interacts directly with ACE2 — the SARS-CoV-2 receptor. Chemokine receptor genes (*CXCR6, CCR9, XCR1*) regulate immune cell trafficking to infected lungs. This haplotype was inherited from Neanderthals and is enriched in Europeans, explaining elevated risk in this population.

- **Chr 9 — ABO locus:** Blood group A individuals showed significantly higher risk of respiratory failure vs blood group O. ABO influences von Willebrand factor levels, endothelial activation, and complement pathways — all relevant to COVID-19 pathophysiology.

- **Genomic inflation (lambda):** A value close to 1.0 confirms population stratification is well controlled, validating the association signals as genuine rather than artefactual.

- **Pathways implicated:** Innate immune response, chemokine receptor signalling, ACE2-dependent viral entry, endothelial biology, and coagulation.

---

# 13. Session Info

```{r session-info}
sessionInfo()
```

---

*Task 6 — GWAS Analysis Complete*  
*Genomics Project Part B | SINES, NUST | April 2026*
